---
title: "Get started with gps2gtfs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with gps2gtfs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`gps2gtfs` turns raw GPS pings from transit vehicles into GTFS-shaped `trips`
and `stop_times` tables. This article runs the whole pipeline on the data that
ships with the package: one day of pings from a bus on route 654 between Kandy
and Digana, Sri Lanka.

```{r setup}
library(gps2gtfs)
library(data.table)
```

## The inputs

The extractor needs three tables:

- **GPS pings**: one row per position report, with a vehicle id, coordinates,
  a timestamp and, ideally, a speed.
- **Terminals**: the points where trips start and end.
- **Stops**: the stops of each direction of the route.

```{r inputs}
head(g2g_data_gps)
g2g_data_terminals
head(g2g_data_stops)
```

The stops are labelled by direction (`Kandy-Digana`, `Digana-Kandy`). Which
label belongs to which terminal cannot be recovered from the data, because
both directions run along the same road. You state it once, as a map from each
label to the terminal its trips *start* from:

```{r direction-map}
direction_map <- c("Kandy-Digana" = "BT01", "Digana-Kandy" = "BT02")
```

## Step 1: clean the pings

`g2g_clean_gps()` normalises the column names, drops pings with a zero
latitude or longitude (the usual sentinel for a missing fix) and pings without
a vehicle id, removes repeated `(vehicle_id, timestamp)`
reports, and sorts each vehicle's pings in time. The extractor calls it
internally, so you only call it yourself to inspect the result:

```{r clean}
gps <- g2g_clean_gps(g2g_data_gps)
nrow(g2g_data_gps)
nrow(gps)
head(gps)
```

The first pings of the day sit at `(0, 0)` and are dropped.

## Step 2: extract trips and stop times

`g2g_extract_trips_and_stop_times()` cuts each vehicle's trajectory into trips
between the terminal buffers, then finds the visits to each stop of the trip's
direction. The buffer radii are in metres.

```{r extract}
result <- g2g_extract_trips_and_stop_times(
  gps_data = g2g_data_gps,
  terminals_data = g2g_data_terminals,
  stops_data = g2g_data_stops,
  terminals_buffer_radius = 50,
  stops_buffer_radius = 30,
  stops_extended_buffer_radius = 50,
  stop_direction_map = direction_map,
  return_trajectory = TRUE
)
```

The bus made six trips, alternating between the two terminals:

```{r trips}
result$trips[, .(trip_id, start_terminal, end_terminal, direction,
                 start_time, end_time, duration_in_mins)]
```

Each stop visit has an arrival, a departure and a dwell time:

```{r stop-times}
head(result$stop_times[, .(trip_id, stop_id, arrival_time, departure_time,
                           dwell_time_in_seconds)], 10)
```

All times are absolute `POSIXct` values in the timezone of the input, not
GTFS clock strings, so a trip that runs past midnight stays unambiguous. These
are inference tables, not GTFS files yet: they carry no `route_id`,
`service_id` or `stop_sequence`. Turning them into a valid feed is the job of
[`gtfsrt2static`](https://e-kotov.github.io/gtfsrt2static/); see
[From raw GPS to a realized GTFS feed](https://e-kotov.github.io/gtfsrt2static/articles/pipeline.html).

## Step 3: check what was kept

The extraction above warned that it lost coverage. Every stage records how
many pings, segments and trips it kept or dropped, so a loss is visible rather
than silent:

```{r diagnostics}
g2g_diagnostics(result)
```

Here the pings dropped at the trip stage fall outside every trip: the bus
standing at a terminal between runs, before its first run and after its last.
Cleaning removed the pings with a zero coordinate and the repeated reports. A large drop at another stage, such
as rows without a trip identity or stops out of buffer range, is the signal to
revisit the inputs or the radii. Silence the warning with
`diagnostics_warn = FALSE` once you have looked.

## Step 4: the shapes actually driven

With `return_trajectory = TRUE` the result also carries the ping-level
trajectory of each trip. `g2g_shapes_from_trips()` turns it into a table shaped
like GTFS `shapes.txt`, with one shape per trip. Unlike planned shapes, these
record the road the vehicle took, detours included:

```{r shapes}
shapes <- g2g_shapes_from_trips(result$trajectory)
head(shapes)
shapes[, .(points = .N, km = round(max(shape_dist_traveled) / 1000, 1)),
       by = shape_id]
```

## Backends

The heavy steps run in one of three backends: compiled Rust (when a Rust
toolchain was available at install time), C++ through Rcpp, or pure R with
`data.table` and `sf`. `backend = "auto"` takes the first available, in that
order. The startup message of each run names the backend used. The output does
not depend on it: same rows, same values, same row order.

```{r backends}
rcpp <- g2g_extract_trips_and_stop_times(
  gps_data = g2g_data_gps,
  terminals_data = g2g_data_terminals,
  stops_data = g2g_data_stops,
  terminals_buffer_radius = 50,
  stops_buffer_radius = 30,
  stops_extended_buffer_radius = 50,
  stop_direction_map = direction_map,
  backend = "rcpp",
  diagnostics_warn = FALSE
)
identical(rcpp$stop_times, result$stop_times)
```

## Where next

- [GTFS-Realtime Vehicle Positions to GTFS](gtfs-realtime.html): reading
  archived feeds with `gtfsrealtime`, deriving stops and terminals from a
  planned feed, and skipping trip inference when the pings carry a `trip_id`.
- [From raw GPS to a realized GTFS feed](https://e-kotov.github.io/gtfsrt2static/articles/pipeline.html):
  assembling these tables into a standard-compliant GTFS feed with
  `gtfsrt2static` and reading it back with `gtfstools`.
