---
title: "GTFS-Realtime Vehicle Positions to GTFS, natively"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GTFS-Realtime Vehicle Positions to GTFS, natively}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
has_gtfsrealtime <- requireNamespace("gtfsrealtime", quietly = TRUE)
```

GTFS-Realtime Vehicle Positions are raw GPS pings with transit annotations —
exactly the data `gps2gtfs` converts into GTFS-shaped `trips` and
`stop_times` tables. Since version 0.2.0, the package's canonical column
names follow the GTFS-Realtime convention (`vehicle_id`, `latitude`,
`longitude`, `timestamp`, `speed`), so the output of
`gtfsrealtime::read_gtfsrt_positions()` feeds the pipeline **directly, with
no adapter**.

```{r setup}
library(gps2gtfs)
library(data.table)
```

## Reading an archived feed

The [gtfsrealtime](https://cran.r-project.org/package=gtfsrealtime) package
parses GTFS-RT protobuf files — single snapshots, gzip/bzip2 files, or a
daily ZIP of archived polls — into a plain data frame:

```{r read-feed, eval = has_gtfsrealtime}
positions <- gtfsrealtime::read_gtfsrt_positions(
  system.file("nyc-vehicle-positions.pb.bz2", package = "gtfsrealtime"),
  timezone = "America/New_York"
)
names(positions)
```

Every column of that data frame is either consumed by `gps2gtfs`
(`vehicle_id`, `latitude`, `longitude`, `timestamp`, `speed`) or passed
through untouched (`trip_id`, `route_id`, `stop_id`, `current_status`,
`bearing`, ...). Pass the timezone the agency operates in: `g2g_clean_gps()`
preserves it, so service days split at local midnight.

## A worked example

A single feed snapshot contains one ping per vehicle — a trajectory needs an
*archive* of polls. For a reproducible example we simulate what a day-long
archive of one bus on a two-terminal route looks like after parsing: two
runs (A → B, then B → A) with GTFS-RT `trip_id` annotations, plus one
unannotated layover ping.

```{r synthetic}
lat <- seq(7.290, 7.320, length.out = 4)
lon <- seq(80.630, 80.660, length.out = 4)
base <- as.POSIXct("2026-06-06 08:00:00", tz = "UTC")

positions <- data.frame(
  vehicle_id = "7482",
  latitude = c(lat, 7.321, rev(lat)),
  longitude = c(lon, 80.661, rev(lon)),
  timestamp = base + c(0, 300, 600, 900, 1200, 1800, 2100, 2400, 2700),
  speed = c(0, 20, 20, 0, 0, 0, 20, 20, 0),
  trip_id = c(rep("CS_1", 4), NA, rep("CS_2", 4))
)
```

## Terminals and stops from a baseline feed

The pipeline needs terminal and stop locations. With a planned (baseline)
static GTFS feed for the same system — a gtfsio/gtfstools-style object or a
path to a GTFS zip — both are derived automatically:

```{r baseline}
baseline <- list(
  trips = data.frame(
    trip_id = c("CS_1", "CS_2"),
    route_id = "R1",
    service_id = "wk"
  ),
  stop_times = data.frame(
    trip_id = rep(c("CS_1", "CS_2"), each = 4),
    stop_id = c("A", "S1", "S2", "B", "B", "S2", "S1", "A"),
    stop_sequence = rep(1:4, 2)
  ),
  stops = data.frame(
    stop_id = c("A", "S1", "S2", "B"),
    stop_name = c("Terminal A", "Stop 1", "Stop 2", "Terminal B"),
    stop_lat = lat,
    stop_lon = lon
  )
)

terminals <- g2g_terminals_from_gtfs(baseline, route_id = "R1")
stops <- g2g_stops_from_gtfs(baseline, route_id = "R1")
terminals
stops
```

Without a baseline, supply the two tables by hand, or — when positions carry
`stop_id` annotations — estimate stop coordinates from the pings themselves
with `g2g_stops_from_positions()`.

## Extracting trips and stop times

Because the positions carry `trip_id` annotations, we can skip spatial trip
inference entirely (`trip_col = "trip_id"`, the *fast path*). Without the
annotation, drop that argument and trips are inferred from terminal-buffer
crossings instead — same output either way.

```{r pipeline}
result <- g2g_extract_trips_and_stop_times(
  gps_data = positions,
  terminals_data = terminals,
  stops_data = stops,
  terminals_buffer_radius = 100,
  stops_buffer_radius = 100,
  stops_extended_buffer_radius = 150,
  trip_col = "trip_id",
  return_trajectory = TRUE
)

result$trips
result$stop_times
```

These are *inference tables*, not finished GTFS files: they use GTFS-style
naming but carry no `route_id`, `service_id`, or `stop_sequence`, and all
times are absolute `POSIXct` values (in the input timezone) rather than GTFS
clock strings — so trips running past midnight stay unambiguous. Assembling
them into a standard-compliant static feed (ID linkage, stop sequencing,
`>24:00:00` time encoding) is the job of a downstream assembler such as
`gtfsrt2static`.

## Shapes: the geometry actually driven

`return_trajectory = TRUE` exposes the ping-level trajectory, which converts
into a `shapes.txt`-shaped table — unlike planned shapes, these traces record
what the vehicle actually drove, including detours:

```{r shapes}
shapes <- g2g_shapes_from_trips(result$trajectory)
head(shapes)
```

## Notes for real archives

* `g2g_clean_gps()` (called internally by the pipeline) deduplicates repeated
  `(vehicle_id, timestamp)` observations — archived feeds re-report unchanged
  positions every poll — and drops pings with missing `vehicle_id` with a
  warning.
* GTFS-RT reports `speed` in meters per second; only `speed == 0` is used
  (dwell-time detection). A missing speed column degrades dwell estimates
  and triggers a warning.
* Data in any other column naming (AVL exports, logger CSVs) is mapped with
  `vehicle_col =` / `time_col =` — no pre-processing step needed.
